---
title: "Projet EPIQUE : analyse des données de phénotypage chez la caille"
author: "CERUTTI Chloé"
date: "26/07/21"
output:
   html_document: 
      toc: TRUE
      toc_float: TRUE
      code_folding: hide
   pdf_document:
      keep_tex: TRUE
---

# Libraries:

```{r}
libraries = c("ggplot2", "dplyr", "gridExtra", "ggpubr", "cowplot", "tidyr", "nlme", "MASS",
              "emmeans", "car", "ComplexHeatmap", "circlize", "viridis", "gridGraphics")

suppressWarnings(suppressMessages(invisible(lapply(libraries, library, character.only = TRUE))))
```

# I. Data reading:

```{r}
# Data table reading:
phenotypes_data = "inputs/Pheno_all.txt"
table = read.table(phenotypes_data, sep="\t", header=TRUE, na.strings="")

# Change the order of treat levels to compare with Control:
df=table
df$batch = as.character(df$batch)
df$treat <- factor(df$treat, levels = c("Control", "IC", "Bisphenol", "Genistein", "5Aza"))
df$line <- factor(df$line, levels = c("S+", "DD", "A", "B"))

# Addition of family link between individuals:
df <- df %>% mutate(family = as.integer(factor(df$father)))
head(df)

# Direct access to variables giving their names:
suppressWarnings(suppressMessages(attach(df)))
names(df) # 1979

# Data table reading: balance between batch 1 and batch 2:
laid_data = "inputs/PonteInd.txt"
table_batch = read.table(laid_data, sep="\t", header=TRUE)
suppressWarnings(suppressMessages(attach(table_batch)))
df_batch=table_batch
df_batch$batch = as.character(df_batch$batch)
df_batch$date = as.character(df_batch$date)
df_batch$date = as.Date(df_batch$date, format = "%d/%m/%Y")

# Select values in time interval:
df_batch = with(df_batch, df_batch[(batch==1 & date >= "2016-11-13" & date <= "2016-12-13") | (batch==2 & date >= "2017-01-03" & date <= "2017-02-02"), ])

# Change the order of treat levels to compare with Control:
df_batch$line <- factor(df_batch$line, levels = c("S+", "DD", "A", "B"))
suppressWarnings(suppressMessages(attach(df_batch)))

df_eq_batch = df_batch %>% group_by(id) %>%  mutate(nb_days = as.numeric(max(date) - min(date)), eggNumber = sum(count)) 
df_gb_id = unique(df_eq_batch[, c("id", "batch", "nb_days", "eggNumber")])
#nrow(df_gb_id) # 617
head(df_gb_id)
```

```{r, fig.width=15, fig.height=7}
# Data table reading: balance between batch 1 and batch 2:
laid_data = "inputs/PonteInd.txt"
table_batch = read.table(laid_data, sep="\t", header=TRUE)
suppressWarnings(suppressMessages(attach(table_batch)))
df_batch=table_batch
df_batch$batch = as.character(df_batch$batch)
df_batch$date = as.character(df_batch$date)
df_batch$date = as.Date(df_batch$date, format = "%d/%m/%Y")

# Select values in time interval:
df_batch = with(df_batch, df_batch[(batch==1 & date >= "2016-11-13" & date <= "2016-12-13") | (batch==2 & date >= "2017-01-03" & date <= "2017-02-02"), ])

# Change the order of treat levels to compare with Control:
df_batch$line <- factor(df_batch$line, levels = c("S+", "DD", "A", "B"))
suppressWarnings(suppressMessages(attach(df_batch)))

df_eq_batch = df_batch %>% group_by(id) %>%  mutate(nb_days = as.numeric(max(date) - min(date)), eggNumber = sum(count)) 
df_gb_id = unique(df_eq_batch[, c("id", "batch", "nb_days", "eggNumber")])
head(df_gb_id)

# Checking the balance between batch 1 and batch 2:
m_nbegg_batch_1 = mean(df_gb_id[,4][df_gb_id[,2]=="1"]) # 17.11628
m_nbegg_batch_2 = mean(df_gb_id[,4][df_gb_id[,2]=="2"]) # 20.35987
#m_nbegg_batch_1 / m_nbegg_batch_2 # 0.8406869 : ratio close to 1 => number of eggs balanced between lot 1 and 2
```

```{r}
nb_f_old_df = length(which(df$batch=="1" & df$sex=="F")) # 447
nb_f_new_df = nrow(df_gb_id) # 617
diff_nb_f = nb_f_new_df - nb_f_old_df # 170

# Replace the egg number:
suppressWarnings(suppressMessages(df$eggNumber <- replace(df$eggNumber, df$ID %in% df_gb_id$id, df_gb_id$eggNumber)))
names(df_gb_id)[names(df_gb_id) == "id"] <- "ID"
df <- left_join(df, df_gb_id, by = "ID")
names(df)[names(df) == "batch.x"] <- "batch"
names(df)[names(df) == "eggNumber.y"] <- "eggNumber"
df = df[,-c(12,15)]
df = df %>% relocate(eggNumber, .after = age1stegg)

# Ratio batch 1 / btach 2 :
mean_eggnb_batch1 = mean(df$eggNumber[which(df$batch == "1")],  na.rm = TRUE) # 19.71617
mean_eggnb_batch2 = mean(df$eggNumber[which(df$batch == "2")],  na.rm = TRUE) # 19.36422
ratio_batch1_2 = mean_eggnb_batch1 / mean_eggnb_batch2 # 1.018176 
suppressWarnings(suppressMessages(attach(df)))

# Distribution du nombre d'oeufs par lot :
hist(dataframe = df, x.var = "eggNumber")
```

```{r}
# Table of age at first egg and egg number:
df_egg = df[, -c(7:10)]

# Removal of individuals with weight number < 3:
df$nbNA <- rowSums(is.na(df[,c("d8weight", "d21weight", "d36weight", "d78weight")]))
nb_ind_df_init = nrow(df) # 1979
suppressWarnings(suppressMessages(attach(df_egg)))
suppressWarnings(suppressMessages(attach(df)))
```

# II. Data distribution:

```{r}
# Histogramme / courbe de densité : weight en fonction du sexe des individus :
hist_legend <- function(dataframe, x.var, color) {
  x.var = rlang::sym(x.var)
    if(color==sex){
      title = "Sex"
      colors=c("coral", "cyan3")
    }else{
      title = "Line"
      colors=c("orange2", "seagreen4", "gold1", "slateblue")
    }
  ggp <- ggplot(dataframe, aes(x= !! x.var, color=color, fill=color)) + theme_bw(base_size = 16) +
    geom_histogram(alpha=0.5) + geom_density(alpha=.2) + guides(fill=guide_legend(title=title)) +
    guides(color=FALSE) + scale_color_manual(values=colors) + scale_fill_manual(values=colors, na.value = "grey54") 
  legend <- get_legend(ggp)
  return(legend)
}

hist_weight <- function(dataframe, x.var, color, x_label) {
  x.var = rlang::sym(x.var)
  y_label <- ""
  color_name = deparse(substitute(color))
  if(x.var == "slope_norm" || x.var == "d8weight"){
    y_label="Counts"
  }
  if(color_name=="sex" || color_name=="dataframe$sex" || color_name=="batch" || color_name=="dataframe$batch"){
    colors=c("coral", "cyan3")
  }else if(color_name=="line" || color_name=="dataframe$line"){
    colors=c("orange2", "seagreen4", "gold1", "slateblue")
  }else{
    colors=c("red1", "darkorange1", "springgreen4", "dodgerblue1", "orchid2")
  }
  ggp <- ggplot(dataframe, aes(x= !! x.var, color=color, fill=color)) +
    geom_histogram(alpha=0.5, position="identity", show.legend = FALSE) +
    labs(x=x_label, y=y_label) + scale_color_manual(values=colors, na.value = "grey54") +
    scale_fill_manual(values=colors, na.value = "grey54") + theme_bw(base_size = 16)
  return(ggp)
}

hist_egg <- function(dataframe, x.var, color, x_label) {
  x.var = rlang::sym(x.var)
  nb_bar = length(table(dataframe[toString(x.var)]))
  color_name = deparse(substitute(color))
  y_label="Counts"
  if(x.var == "eggNumber" || x.var == "age1stegg"){
    y_label <- ""
  }
  if(color_name=="sex" || color_name=="dataframe$sex"){
    colors="coral"
  }else if(color_name=="line" || color_name=="dataframe$line"){
    colors=c("orange2", "seagreen4", "gold1", "slateblue")
  }else if(color_name=="batch" || color_name=="dataframe$batch"){
    colors=c("coral", "cyan3")
  }else{
    colors=c("red1", "darkorange1", "springgreen4", "dodgerblue1", "orchid2")
  }
  ggp <- ggplot(dataframe, aes(x= !! x.var, color=color, fill=color)) +
    geom_histogram(bins = nb_bar, alpha=0.5, position="identity", show.legend = FALSE) +
    scale_color_manual(values=colors) + scale_fill_manual(values=colors) +
    labs(x=x_label, y=y_label) + theme_bw(base_size = 16)
  return(ggp)
}
```

### a. Histograms by sex: 

```{r,fig.width=15, fig.height=7}
suppressWarnings(suppressMessages({legend_a = hist_legend(dataframe = df, x.var = "d8weight", sex) 
h_d8_sex = hist_weight(dataframe = df, x.var = "d8weight", sex, "Weight at 8 days") 
h_d21_sex = hist_weight(dataframe = df, x.var = "d21weight", sex, "Weight at 21 days") 
h_d36_sex = hist_weight(dataframe = df, x.var = "d36weight", sex, "Weight at 36 days") 
h_d78_sex = hist_weight(dataframe = df, x.var = "d78weight", sex, "Weight at 78 days") 
h_age1stegg_sex = hist_egg(dataframe = df, x.var = "age1stegg", sex, "Age at first egg laid") 
h_eggNumber_sex = hist_egg(dataframe = df, x.var = "eggNumber", sex, "Egg number laid")

# pdf(file = "figures/supp_data_1_a.pdf", width=15, height=7)
grid.arrange(h_d8_sex, h_d21_sex, h_d36_sex, h_d78_sex,
             h_age1stegg_sex, h_eggNumber_sex, legend_a,
             ncol = 4, nrow = 2)
}))
# dev.off()
```

### b. Histograms by line: 

```{r,fig.width=15, fig.height=7}
suppressWarnings(suppressMessages({legend_b = hist_legend(dataframe = df, x.var = "d8weight", line)
h_d8_line = hist_weight(dataframe = df, x.var = "d8weight", line, "Weight at 8 days")
h_d21_line = hist_weight(dataframe = df, x.var = "d21weight", line, "Weight at 21 days")
h_d36_line = hist_weight(dataframe = df, x.var = "d36weight", line,"Weight at 36 days")
h_d78_line = hist_weight(dataframe = df, x.var = "d78weight", line, "Weight at 78 days")
h_age1stegg_line = hist_egg(dataframe = df, x.var = "age1stegg", line, "Age at first egg laid")
h_eggNumber_line = hist_egg(dataframe = df, x.var = "eggNumber", line, "Egg number laid")

# pdf(file = "figures/supp_data_1_b.pdf", width=15, height=7)
grid.arrange(h_d8_line, h_d21_line, h_d36_line, h_d78_line,
             h_age1stegg_line, h_eggNumber_line, legend_b,
             ncol = 4, nrow = 2)
}))
# dev.off()
```

```{r,fig.width=15, fig.height=14}
suppressWarnings(suppressMessages({
first_col = plot_grid(h_d8_line, h_d21_line, h_d36_line, h_d78_line, 
            h_age1stegg_line, h_eggNumber_line, legend_b, ncol=4, nrow=2, labels = "a", label_size = 20)
second_col = plot_grid(h_d8_sex, h_d21_sex, h_d36_sex, h_d78_sex, 
             h_age1stegg_sex, h_eggNumber_sex, legend_a, ncol=4, nrow=2, labels = "b", label_size = 20)
perfect = plot_grid(first_col, second_col, ncol=1, nrow=2, rel_widths = c(15,15), rel_heights = c(10,10))

# pdf(file = "Supplementary_data_1.pdf", width=15, height=14)
suppressWarnings(suppressMessages(perfect))}))
# dev.off()
```

```{r}
# Mean age1stegg and standard deviation (sd) for each line : 
df_mean_sd_age1stegg = df %>% group_by(line) %>% summarise(mean = round(mean(na.omit(age1stegg)), digits = 1), sd = round(sd(na.omit(age1stegg)), digits = 2))
df_mean_sd_age1stegg

# Mean eggNumber and standard deviation (sd) for each line : 
df_mean_sd_eggnumber = df %>% group_by(line) %>% summarise(mean = round(mean(na.omit(eggNumber)), digits = 1), sd = round(sd(na.omit(eggNumber)), digits = 2))
df_mean_sd_eggnumber
```

# III. Statistical analysis:

## 1. Independent variables (aov (+) + TukeyHSD) :

```{r}
create_table_with_sign <- function(data, type) {
  if(type=="treat"){data_sign = data.frame(data$treat)}
  if(type=="sex:treat"){data_sign = data.frame(data$`sex:treat`)}
  if(type=="batch:treat"){data_sign = data.frame(data$`batch:treat`)}
  if(type=="line:sex"){data_sign = data.frame(data$`line:sex`)}
  if(type=="treat:sex"){data_sign = data.frame(data$`treat:sex`)}
  if(type=="line:treat"){data_sign = data.frame(data$`line:treat`)}
  if(type=="line:treat:sex"){data_sign = data.frame(data$`line:treat:sex`)}
  data_sign$sign <- ifelse(data_sign$p.adj < 0.001, "***",
                                  ifelse(data_sign$p.adj < 0.01, "**",
                                         ifelse(data_sign$p.adj < 0.05, "*",
                                                ifelse(data_sign$p.adj < 0.1, ".","ns"))))
  return(data_sign) }

create_short_sign_table = function(data) {
  final_data = data[data[, "sign"] != "ns",]
  return(final_data) }

create_sign_tukeyHSD_table <- function(data, type){
  tab = create_table_with_sign(data=data, type)
  final_tab = create_short_sign_table(data=tab)
  return(final_tab) }
```

## 2. Model selection:

```{r,fig.width=15, fig.height=13}
df = df[,c("ID", "father", "mother", "sex", "line", "treat", "d8weight", "d21weight", "d36weight", "d78weight", "age1stegg", "eggNumber", "batch", "family")]

add_weight <- function(dataframe) {
  dataframe = dataframe %>% gather(age, weight, c("d8weight", "d21weight", "d36weight", "d78weight"))
  dataframe = dataframe[order(dataframe$ID),]
  dataframe["age"][dataframe["age"]=="d8weight"] <- 8
  dataframe["age"][dataframe["age"]=="d21weight"] <- 21
  dataframe["age"][dataframe["age"]=="d36weight"] <- 36
  dataframe["age"][dataframe["age"]=="d78weight"] <- 78
  rownames(dataframe) <- NULL
  return(dataframe)
}

df = add_weight(dataframe = df)
df$treat <- factor(df$treat, levels = c("IC", "Control", "Bisphenol", "Genistein", "5Aza"))
df_egg$treat <- factor(df_egg$treat, levels = c("IC", "Control", "Bisphenol", "Genistein", "5Aza"))

# Convert class age to numeric:
df$age <- as.numeric(df$age)

# Mean weight and standard deviation (sd) for each line : 
df_mean_weight_age = df %>% group_by(line) %>% summarise(mean_weight = round(mean(na.omit(weight)), digits = 1), 
                                                         sd_weight = round(sd(na.omit(weight)), digits = 2))
df_mean_weight_age
```

### a. Weight ( 1 + age + age²) :

#### Lme and model selection with stepAIC (random effect 1|ID) :

```{r}
lmemod_weight = lme(weight ~ 1 + (treat * age * sex * line * batch) + (I(age^2) * treat * sex * line * batch),  
                      data=df[,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")],
                      random=~1|ID, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))
mod_weight = stepAIC(lmemod_weight) 
mod_weight
```

#### Lme on model selected by stepAIC :

```{r}
# Random effect family/ID:
lmemod_weight_selected_tmp = lme(weight ~ treat + age + sex + line + batch + I(age^2) + treat:sex + age:sex + treat:line + age:line + sex:line + treat:batch + age:batch + sex:batch + line:batch + treat:I(age^2) + sex:I(age^2) + line:I(age^2) + batch:I(age^2) + treat:sex:line + age:sex:line + treat:sex:batch + treat:line:batch + sex:line:batch + treat:sex:I(age^2) + treat:line:I(age^2) + sex:line:I(age^2) + treat:batch:I(age^2) + sex:batch:I(age^2) + line:batch:I(age^2),
                               data=df[,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")],
                               random=~(1+age)|family/ID, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))

lmemod_weight_selected = lme(weight ~ treat + age + sex + line + batch + I(age^2) + treat:sex + age:sex + age:line + sex:line + age:batch + sex:batch + treat:I(age^2) + sex:I(age^2) + line:I(age^2) + batch:I(age^2) + age:sex:line + treat:sex:I(age^2) + sex:line:I(age^2) + sex:batch:I(age^2), 
                               data=df[,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")],
                               random=~(1+age)|family/ID, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))

tuk_weight = emmeans(lmemod_weight_selected, pairwise ~ treat, adjust="tukey", data=df)
```

#### Anova:

```{r}
Anova(lmemod_weight_selected_tmp, type=3) 
```

```{r}
Anova(lmemod_weight_selected, type=3) 
```

#### Summary:

```{r}
printCoefmat(summary(lmemod_weight_selected)$tTable)
```

### b. Age1stegg:

```{r}
suppressWarnings(suppressMessages(attach(df_egg)))

lmemod_age1stegg = lme(age1stegg ~ treat * line * batch, random=~1|family, data=df_egg[,c("ID", "line", "batch", "treat", "family", "age1stegg")], na.action=na.exclude, method="ML")
mod_age1stegg = stepAIC(lmemod_age1stegg)
mod_age1stegg
tuk_age1stegg = emmeans(mod_age1stegg, pairwise ~ treat, adjust="tukey", data=df_egg)
```

#### Anova:

```{r}
Anova(mod_age1stegg, type=3)
```

#### Summary:

```{r}
printCoefmat(summary(mod_age1stegg)$tTable)
```

### c. EggNumber:

```{r}
lmemod_eggnb = lme(eggNumber ~ treat * line * batch, random=~1|family, data=df_egg[,c("ID", "line", "batch", "treat", "family", "eggNumber")], na.action=na.exclude, method="ML")
mod_eggnb = stepAIC(lmemod_eggnb)
mod_eggnb
```

#### Anova:

```{r}
Anova(mod_eggnb, type=3)
```
#### Summary:

```{r}
printCoefmat(summary(mod_eggnb)$tTable)
```

#### Treatment:sex pairwise comparison with emmeans (TukeyHSD):

#### P-values intercept, age, age²:

```{r}
suppressWarnings(suppressMessages(attach(df)))
emmeans(lmemod_weight_selected, pairwise ~ treat*sex | age, adjust="tukey", at = list(age=c(0,32.713)))
```

#### P-values weight at 8 days :

```{r}
# data: 
df_8 = df[df["age"]==8,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")]

# lme:
lmemod_weight_d8 = lme(weight ~ treat * line * batch * sex, random=~1|family,
                       data=df_8, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))

# stepAIC:
mod_d8 = stepAIC(lmemod_weight_d8)

# tukey:
tuk_emm_d8 = emmeans(mod_d8, pairwise ~ treat * sex, adjust="tukey", data=df_8)
```

```{r}
mod_d8
Anova(mod_d8, type=3)
printCoefmat(summary(mod_d8)$tTable)
```

#### P-values weight at 21 days : 

```{r}
df_21 = df[df["age"]==21,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")]

lmemod_weight_d21 = lme(weight ~ treat * line * batch * sex, random=~1|family,
                        data=df_21, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))

mod_d21 = stepAIC(lmemod_weight_d21)

tuk_emm_d21 = emmeans(mod_d21, pairwise ~ treat * sex, adjust="tukey", data=df_21)
```

```{r}
mod_d21
Anova(mod_d21, type=3)
printCoefmat(summary(mod_d21)$tTable)
```

#### P-values weight at 36 days : 

```{r}
df_36 = df[df["age"]==36,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")]

lmemod_weight_d36 = lme(weight ~ treat * line * batch * sex, random=~1|family,
                        data=df_36, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))

mod_d36 = stepAIC(lmemod_weight_d36)

tuk_emm_d36  = emmeans(mod_d36, pairwise ~ treat * sex, adjust="tukey", data=df_36)
```

```{r}
mod_d36
Anova(mod_d36, type=3)
printCoefmat(summary(mod_d36)$tTable)
```

#### P-values weight at 78 days :

```{r}
df_78 = df[df["age"]==78,c("ID", "sex", "line", "batch", "treat", "family", "weight", "age")]

lmemod_weight_d78 = lme(weight ~ treat * line * batch * sex, random=~1|family,
                        data=df_78, na.action=na.exclude, method="ML", control=lmeControl(opt='optim'))

mod_d78 = stepAIC(lmemod_weight_d78)

tuk_emm_d78 = emmeans(mod_d78, pairwise ~ treat * sex, adjust="tukey", data=df_78)
```

```{r}
mod_d78
Anova(mod_d78, type=3)
printCoefmat(summary(mod_d78)$tTable)
```

## 3. Heatmaps of Tukey HSD results: 

```{r}
heatmap2 <- function(data, row_legend, top_title) {
  data = data[,  c("p.adj", "sextreat1", "sextreat2")]
  test1 <- data %>% mutate(temp=sextreat1) %>% mutate(sextreat1=sextreat2) %>% mutate(sextreat2=temp) %>% mutate(temp=NULL)
  data <- data %>% mutate(sextreat1=sextreat1)
  result <- rbind(data, test1)
  result <- result[order(result$p.adj),]
  mat_tmp = spread(result, sextreat1, p.adj)
  rownames(mat_tmp) <- mat_tmp$sextreat2
  mat_tmp <- mat_tmp[, -1]
  new_order <- c("F:Control", "F:IC", "F:Bisphenol", "F:Genistein", "F:5Aza", "M:Control", "M:IC", "M:Bisphenol", "M:Genistein", "M:5Aza")
  mat_tmp <- mat_tmp[new_order, new_order]
  mat = as.matrix(mat_tmp)
  class(mat) <- "numeric"
  mat2 <- mat
  ha_top = HeatmapAnnotation(empty = anno_empty(border = FALSE, height = unit(1, "mm")),
                             foo = anno_block(gp = gpar(fill = c("coral", "cyan3")), labels = c("F", "M")))
  ha_left = rowAnnotation(foo = anno_block(gp = gpar(fill = c("coral", "cyan3")), labels = c("F", "M")))
  split_top = rep(1:2, each = 5)
  split_left = rep(1:2, each = 5)
  # col_fun = colorRamp2(0:10, viridis(11))
  col_fun = colorRamp2(0:8, viridis(9))
  col_fun(seq(-3, 3))

  if (row_legend==T) {
     Heatmap(mat2, name = "-log10(p-value)", column_title = top_title, na_col = "white",
          column_title_gp = gpar(fill = "azure4", col = "white", border = "black", fontface = "bold"),
          cluster_columns = FALSE, cluster_rows = FALSE, col = col_fun,
          cluster_row_slices = FALSE, cluster_column_slices = FALSE,
          column_split = split_top, row_split = split_left,
          top_annotation = ha_top, left_annotation = ha_left,
          row_title = NULL, column_names_rot = 45, border = TRUE)
  }else{
    Heatmap(mat2, name = "-log10(p-value)",
          cluster_columns = FALSE, cluster_rows = FALSE, col = col_fun, na_col = "white",
          column_title = top_title,
          column_title_gp = gpar(fill = "azure4", col = "white", border = "black", fontface = "bold"),
          cluster_row_slices = FALSE, cluster_column_slices = FALSE,
          column_split = split_top, row_split = split_left,
          top_annotation = ha_top, row_title = NULL, column_names_rot = 45, border = TRUE)
  }
}

tukey_pvalues_to_df <- function(dataframe) {
  new_df = dataframe$contrasts %>% as.data.frame() %>%
                separate(contrast, c("treat1", "sex1", "sep", "treat2", "sex2"), " ") %>%
                unite("sextreat1", c(sex1,treat1), sep=":", remove=FALSE) %>%
                unite("sextreat2", c(sex2,treat2), sep=":", remove=FALSE) %>%
                dplyr::select("sextreat1", "sextreat2", "p.value") %>% 
                dplyr::rename(p.adj=p.value)
  return(new_df)
} 

df_with_sign <- function(dataframe){
  dataframe$sign <- ifelse(dataframe$p.value < 0.001, "***",
                                  ifelse(dataframe$p.value < 0.01, "**",
                                         ifelse(dataframe$p.value < 0.05, "*",
                                                ifelse(dataframe$p.value < 0.1, ".","ns"))))
  return(dataframe)
}
```

```{r}
df_tuk_emm_d8 = tukey_pvalues_to_df(dataframe = tuk_emm_d8)
df_tuk_emm_d21 = tukey_pvalues_to_df(dataframe = tuk_emm_d21)
df_tuk_emm_d36 = tukey_pvalues_to_df(dataframe = tuk_emm_d36)
df_tuk_emm_d78 = tukey_pvalues_to_df(dataframe = tuk_emm_d78)

df_with_sign(data.frame(tuk_emm_d8$contrasts))
df_with_sign(data.frame(tuk_emm_d21$contrasts))
df_with_sign(data.frame(tuk_emm_d36$contrasts))
df_with_sign(data.frame(tuk_emm_d78$contrasts))

write.table(df_with_sign(data.frame(tuk_emm_d8$contrasts)), file='d8weight.tsv', quote=FALSE, sep='\t')
write.table(df_with_sign(data.frame(tuk_emm_d21$contrasts)), file='d21weight.tsv', quote=FALSE, sep='\t')
write.table(df_with_sign(data.frame(tuk_emm_d36$contrasts)), file='d36weight.tsv', quote=FALSE, sep='\t')
write.table(df_with_sign(data.frame(tuk_emm_d78$contrasts)), file='d78weight.tsv', quote=FALSE, sep='\t')
```

### a. Heatmaps of Tukey HSD p-values: 

```{r,fig.width=18, fig.height=5.5}
# -log10(p_values):
df_tuk_emm_d8$p.adj <- -log10(df_tuk_emm_d8$p.adj + 1e-12)
df_tuk_emm_d21$p.adj <- -log10(df_tuk_emm_d21$p.adj + 1e-12)
df_tuk_emm_d36$p.adj <- -log10(df_tuk_emm_d36$p.adj + 1e-12)
df_tuk_emm_d78$p.adj <- -log10(df_tuk_emm_d78$p.adj + 1e-12)

hm_d8 = heatmap2(data = df_tuk_emm_d8, T, "Weight at 8 days")
hm_d21 = heatmap2(data = df_tuk_emm_d21, F, "Weight at 21 days")
hm_d36 = heatmap2(data = df_tuk_emm_d36, F, "Weight at 36 days")
hm_d78 = heatmap2(data = df_tuk_emm_d78, F, "Weight at 78 days")

hm_list = hm_d8 + hm_d21 + hm_d36 + hm_d78

hp_random_family_ID_family = draw(hm_list, merge_legend = TRUE, legend_title_gp = gpar(fontsize = 14))

# pdf(file = "figure_2_a.pdf", width=18, height=5.5)
# hp_random_family_ID_family
# dev.off()
```

### b. Heatmaps of Tukey HSD estimate values: 

```{r}
heatmap3 <- function(data, row_legend, top_title) {
  data = data[,  c("estimate", "sextreat1", "sextreat2")]
  test1 <- data %>% mutate(temp=sextreat1) %>% mutate(sextreat1=sextreat2) %>% mutate(sextreat2=temp) %>% mutate(temp=NULL)
  data <- data %>% mutate(sextreat1=sextreat1)
  result <- rbind(data, test1)
  result <- result[order(result$estimate),]
  mat_tmp = spread(result, sextreat1, estimate)
  rownames(mat_tmp) <- mat_tmp$sextreat2
  mat_tmp <- mat_tmp[, -1]
  new_order <- c("F:Control", "F:IC", "F:Bisphenol", "F:Genistein", "F:5Aza", "M:Control", "M:IC", "M:Bisphenol", "M:Genistein", "M:5Aza")
  mat_tmp <- mat_tmp[new_order, new_order]
  mat = as.matrix(mat_tmp)
  class(mat) <- "numeric"
  mat2 <- mat
  ha_top = HeatmapAnnotation(empty = anno_empty(border = FALSE, height = unit(1, "mm")),
                             foo = anno_block(gp = gpar(fill = c("coral", "cyan3")), labels = c("F", "M")))
  ha_left = rowAnnotation(foo = anno_block(gp = gpar(fill = c("coral", "cyan3")), labels = c("F", "M")))
  split_top = rep(1:2, each = 5)
  split_left = rep(1:2, each = 5)
  col_fun=colorRamp2(c(-20, 0, 20), c("deeppink3", "white", "darkgreen"))
  # col_fun=colorRamp2(c(-40, 0, 40), c("deeppink3", "white", "darkgreen"))
  col_fun(seq(-3, 3))

  if (row_legend==T) {
     Heatmap(mat2, name = "estimate", column_title = top_title, na_col = "white",
          column_title_gp = gpar(fill = "azure4", col = "white", border = "black", fontface = "bold"),
          # cluster_columns = FALSE, cluster_rows = FALSE, col = colorRamp2(myBreaks, col_fun),
          cluster_columns = FALSE, cluster_rows = FALSE, col = col_fun,
          cluster_row_slices = FALSE, cluster_column_slices = FALSE,
          column_split = split_top, row_split = split_left,
          top_annotation = ha_top, left_annotation = ha_left,
          row_title = NULL, column_names_rot = 45, border = TRUE)
  }else{
    Heatmap(mat2, name = "estimate",
          cluster_columns = FALSE, cluster_rows = FALSE, na_col = "white",
          column_title = top_title,
          column_title_gp = gpar(fill = "azure4", col = "white", border = "black", fontface = "bold"),
          # cluster_row_slices = FALSE, cluster_column_slices = FALSE,  col = colorRamp2(myBreaks, col_fun),
          cluster_row_slices = FALSE, cluster_column_slices = FALSE,  col = col_fun,
          column_split = split_top, row_split = split_left,
          top_annotation = ha_top, row_title = NULL, column_names_rot = 45, border = TRUE)
  }
}
```

```{r}
tukey_estimate_to_df <- function(dataframe) {
  new_df = dataframe$contrasts %>% as.data.frame() %>%
                separate(contrast, c("treat1", "sex1", "sep", "treat2", "sex2"), " ") %>%
                unite("sextreat1", c(sex1,treat1), sep=":", remove=FALSE) %>%
                unite("sextreat2", c(sex2,treat2), sep=":", remove=FALSE) %>%
                select("sextreat1", "sextreat2", "estimate")
  return(new_df)
}
```

```{r}
df_tuk_emm_d8_e = tukey_estimate_to_df(dataframe = tuk_emm_d8)
df_tuk_emm_d8_e
df_tuk_emm_d21_e = tukey_estimate_to_df(dataframe = tuk_emm_d21)
df_tuk_emm_d21_e
df_tuk_emm_d36_e = tukey_estimate_to_df(dataframe = tuk_emm_d36)
df_tuk_emm_d36_e
df_tuk_emm_d78_e = tukey_estimate_to_df(dataframe = tuk_emm_d78)
df_tuk_emm_d78_e
```

```{r,fig.width=18, fig.height=5.5}
hm_d8_e = heatmap3(data = df_tuk_emm_d8_e, T, "Weight at 8 days")
hm_d21_e = heatmap3(data = df_tuk_emm_d21_e, F, "Weight at 21 days")
hm_d36_e = heatmap3(data = df_tuk_emm_d36_e, F, "Weight at 36 days")
hm_d78_e = heatmap3(data = df_tuk_emm_d78_e, F, "Weight at 78 days")

hm_list_e = hm_d8_e + hm_d21_e + hm_d36_e + hm_d78_e

hp_random_family_ID_family_e = draw(hm_list_e, merge_legend = TRUE, legend_title_gp = gpar(fontsize = 14))

# pdf(file = "figure_2_b.pdf", width=18, height=5.5)
# hp_random_family_ID_family_e
# dev.off()
```

# IV. Weight evolution according to the age, line and sex: 

```{r,fig.width=5, fig.height=4}
df_summary <- na.omit(df[,c("ID", "sex", "line", "weight", "age")]) %>% group_by(line, sex, age) %>% summarise(sd = sd(weight, na.rm = TRUE), weight = mean(weight))
df_summary
df_summary$line <- factor(df_summary$line, levels = c("A", "B", "S+", "DD"))

plot_age_line_sex = ggplot(df_summary, aes(age, weight, color=line, linetype = sex)) +
  geom_line() +
  scale_color_manual(values=c("gold1", "slateblue", "orange2", "seagreen4")) +
  geom_errorbar(aes(ymin = weight-sd, ymax = weight+sd), width = 0.8) +
  geom_point(size = 1.5) + 
  theme_bw() +
  theme(text = element_text(size = 12)) + 
  labs(x = "Age (days)", y="Weight (g)")

# pdf(file = "figure_1.pdf", width=5, height=4)
plot_age_line_sex
# dev.off()
```
